---
title: "NHL Stanley Cup Playoff Predictions"
author: "Ilari Scheinin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

Previous: [README](https://github.com/ilarischeinin/stanley)  
Next: [2. Scrape raw data](scrape.html)

1. Canadian programming
-----------------------

I have played ice hockey for more than 25 years, and I feel passionate about the sport. I feel that nowhere else, not even in my native Finland, do people truly *get* ice hockey in the same way as Canadians do. Therefore, as a sign of appreciation, I have written this project following conventions of
[Canadian programming](http://programmers.stackexchange.com/a/5576). For
example, instead of function `is.numeric()`, I am using `numeric.eh()`.

These function definitions are stored in file `canadian.R`,
which is `source()`d to the main analysis scripts.

```{r, eval=FALSE}
source("canadian.R")
```

```{r, echo=FALSE, comment=""}
cat(readLines("canadian.R"), sep="\n")
```

Next: [2. Scrape raw data](scrape.html)  
Previous: [README](https://github.com/ilarischeinin/stanley)
