---
title: "NHL Stanley Cup Playoff Predictions"
author: "Ilari Scheinin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

Previous: [3. Process data](process.html)  
Next: [5. Make predictions](predict.html)

4. Train models
---------------

Load packages and the Canadian programming library. Please note that
`caret` automatically loads `plyr`. As I'm also using `dplyr`, it's crucial
that `dplyr` is loaded after `plyr`. Doing it the other way around causes
problems. Also define two CPU cores to be used for model training.

```{r}
suppressMessages({
  library(plyr)
  library(dplyr)
  library(caret)
  library(doMC)
})
source("canadian.R")
registerDoMC(cores=2)
```

Function to detect home/away team and winner.

```{r}
team <- function(x, tied.eh=c("silent", "warn", "stop")) {
  counts <- sort(table(x), decreasing=TRUE)
  if ((length(counts) == 1) || (counts[1] > counts[2]))
    return(names(counts)[1])
  tied.eh <- match.arg(tied.eh)
  if (tied.eh == "error")
    stop("I'm sorry, but I couldn't figure out the winner: ",
      paste(x, collapse=", "))
  if (tied.eh == "warn")
    warning("I couldn't figure out the winner: ", paste(x, collapse=", "))
  x[1]
}
```

Function to add game statistis.

```{r}
add_stats <- function(games, gamestats, which=c("both", "single", "overall")) {
  which <- match.arg(which)
  if (which == "overall") {
    away <- left_join(games, gamestats[["overall"]],
      by=c("season", awayteam="team"))
    home <- left_join(games, gamestats[["overall"]],
      by=c("season", hometeam="team"))
  } else {
    away <- left_join(games, gamestats[["away"]],
      by=c("season", awayteam="team"))
    home <- left_join(games, gamestats[["home"]],
      by=c("season", hometeam="team"))
  }

  if (which == "both") {
    away2 <- left_join(games, gamestats[["home"]],
      by=c("season", awayteam="team"))
    home2 <- left_join(games, gamestats[["away"]],
      by=c("season", hometeam="team"))
  }

  games$goals <- away$goals - home$goals
  games$shots <- away$shots - home$shots
  games$faceoffs <- away$faceoffs - home$faceoffs
  games$penalties <- away$penalties - home$penalties
  games$pp <- away$pp - home$pk
  games$pk <- away$pk - home$pp

  if (which == "both") {
    games$goals2 <- away2$goals - home2$goals
    games$shots2 <- away2$shots - home2$shots
    games$faceoffs2 <- away2$faceoffs - home2$faceoffs
    games$penalties2 <- away2$penalties - home2$penalties
    games$pp2 <- away2$pp - home2$pk
    games$pk2 <- away2$pk - home2$pp
  }

  games
}
```

Load and process.

```{r}
load(file.path("source-data", "nhlscrapr-core.RData"))
rm(list=c("roster.master", "roster.unique"))

gamestats <- readRDS("processed.rds")

games <- tbl_df(games)
games <- games %>%
  filter(status != 0, session == "Playoffs", season != "20142015") %>%
  mutate(awayscore=as.integer(awayscore), homescore=as.integer(homescore))

playoffs <- games %>%
  mutate(winner=ifelse(awayscore > homescore, awayteam, hometeam)) %>%
  group_by(season, round=substring(gcode, 3, 3),
    series=substring(gcode, 4, 4)) %>%
  summarise(awayteam=team(awayteam), hometeam=team(hometeam),
    winner=team(winner, tied.eh="stop")) %>%
  select(season, round, series, awayteam, hometeam, winner) %>%
  mutate(winner=ifelse(awayteam == winner, "away", "home"))

playoffs <- add_stats(playoffs, gamestats)

playoffs <- playoffs %>%
  ungroup() %>%
  mutate(season=as.integer(substring(season, 5, 8))) %>%
  select(-season, -round, -series, -awayteam, -hometeam)

playoffs$winner = as.factor(playoffs$winner)
```

```{r}
# settings
method <- "repeatedcv"
number <- 10
repeats <- 10
preProcess <- c("center", "scale")
tuneLength <- 5
metric <- "Accuracy"
maxVariables <- 5
seeds <- vector(mode="list", length=repeats*number+1)
for (i in seq_along(seeds))
  seeds[[i]] <- (1000*i+1):(1000*i+1+tuneLength^maxVariables)

fitControl <- trainControl(method=method, number=number, repeats=repeats,
  seeds=seeds)

model <- function(method) {
  message("Training model: ", method, "...", appendLF=FALSE)
  set.seed(7474505)
  suppressMessages({ captured <- capture.output({
    fit <- train(winner ~ ., data=playoffs,
      method=method, trControl=fitControl, preProcess=preProcess,
      metric=metric, tuneLength=tuneLength)
  })})
  message()
  fit
}

methods <- c("glm", "lda", "nnet", "rf", "svmLinear")
models <- lapply(methods, model)
names(models) <- methods

saveRDS(models, "models.rds")

resamps <- resamples(models)
summary(resamps)

trellis.par.set(caretTheme())
bwplot(resamps, layout=c(2, 1))
dotplot(resamps, metric="Accuracy")
dotplot(resamps, metric="Kappa")
```

Next: [5. Make predictions](predict.html)  
Previous: [3. Process data](process.html)
