---
title: "NHL Stanley Cup Playoff Predictions"
author: "Ilari Scheinin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

Previous: [4. Train models](model.html)  
Next: [README](https://github.com/ilarischeinin/stanley)

5. Make predictions
-------------------

In the previous step five different types of models (glm, lda, nnet, rf,
svmLinear) were fitted to the training data of Stanley Cup playoffs from 2003
to 2014. Now I want to use those models to make predictions for this year.

Start by loading packages and the Canadian programming library:

```{r, echo=FALSE}
options(width=120)
```

```{r}
suppressMessages({
  library(plyr)
  library(dplyr)
  library(caret)
})
source("canadian.R")
```

Load the models and summary statistics.

```{r}
gamestats <- readRDS("processed.rds")
models <- readRDS("models.rds")
```

Any one of the five models could be used to make predictions. 

the [Kappa statistic](https://en.wikipedia.org/wiki/Cohen%27s_kappa)

resamps <- resamples(models)
summary(resamps)

trellis.par.set(caretTheme())
bwplot(resamps, layout=c(2, 1))
dotplot(resamps, metric="Accuracy")
dotplot(resamps, metric="Kappa")


# accuracy <- function(fit) { max(fit$result[, fit$metric]) }
# accuracies <- sapply(models, accuracy)
#
# kappa <- function(fit) { max(fit$result[, "Kappa"]) }
# kappas <- sapply(models, kappa)
#
# fit <- models[[which.max(kappas)]]
```

Include the same `add_stats()` function that was used when training the models.



```{r}
add_stats <- function(games, gamestats, which=c("both", "single", "overall")) {
  which <- match.arg(which)
  if (which == "overall") {
    away <- left_join(games, gamestats[["overall"]],
      by=c("season", awayteam="team"))
    home <- left_join(games, gamestats[["overall"]],
      by=c("season", hometeam="team"))
  } else {
    away <- left_join(games, gamestats[["away"]],
      by=c("season", awayteam="team"))
    home <- left_join(games, gamestats[["home"]],
      by=c("season", hometeam="team"))
  }

  if (which == "both") {
    away2 <- left_join(games, gamestats[["home"]],
      by=c("season", awayteam="team"))
    home2 <- left_join(games, gamestats[["away"]],
      by=c("season", hometeam="team"))
  }

  games$goals <- away$goals - home$goals
  games$shots <- away$shots - home$shots
  games$faceoffs <- away$faceoffs - home$faceoffs
  games$penalties <- away$penalties - home$penalties
  games$pp <- away$pp - home$pk
  games$pk <- away$pk - home$pp

  if (which == "both") {
    games$goals2 <- away2$goals - home2$goals
    games$shots2 <- away2$shots - home2$shots
    games$faceoffs2 <- away2$faceoffs - home2$faceoffs
    games$penalties2 <- away2$penalties - home2$penalties
    games$pp2 <- away2$pp - home2$pk
    games$pk2 <- away2$pk - home2$pp
  }

  games
}
```

```{r}

winners <- function(games, models, gamestats) {
  suppressMessages({ captured <- capture.output({
    predictions <- sapply(models, predict, newdata=add_stats(games, gamestats))
  })})
  if (nrow(games) > 1) {
    concensus <- apply(predictions, 1, function(x)
      names(sort(table(x), decreasing=TRUE))[1])
  } else {
    concensus <- names(sort(table(predictions), decreasing=TRUE))[1]
  }
  result <- games$hometeam
  result[concensus == "away"] <- games$awayteam[concensus == "away"]
  result
}
```

Define the playoff games for round 1, and all possible game combinations for
subsequent rounds.

```{r}
round1 <- data_frame(season="20142015",
  awayteam=c("PIT", "OTT", "DET", "NYI", "WPG", "MIN", "CHI", "CGY"),
  hometeam=c("NYR", "MTL", "T.B", "WSH", "ANA", "STL", "NSH", "VAN"))

round2 <- data_frame(season="20142015",
  awayteam=c("WSH", "NYI", "T.B", "DET", "OTT", "OTT", "PIT", "PIT",
    "VAN", "CGY", "NSH", "CHI", "MIN", "MIN", "WPG", "CGY"),
  hometeam=c("NYR", "NYR", "MTL", "MTL", "T.B", "DET", "WSH", "NYI",
    "ANA", "ANA", "STL", "STL", "NSH", "CHI", "VAN", "WPG"))

round3 <- data_frame(season="20142015",
  awayteam=c(
    "MTL", "T.B", "DET", "OTT",
    "WSH", "NYI", "PIT",
    "WSH", "NYI", "PIT",
    "DET", "OTT",
    "DET", "OTT",
    "PIT",
    "PIT",
    "STL", "NSH", "CHI", "MIN",
    "VAN", "WPG", "CGY",
    "VAN", "WPG", "CGY",
    "VAN", "WPG", "CGY",
    "MIN",
    "WPG", "CGY"),
  hometeam=c(
    "NYR", "NYR", "NYR", "NYR",
    "MTL", "MTL", "MTL",
    "T.B", "T.B", "T.B",
    "WSH", "WSH",
    "NYI", "NYI",
    "DET",
    "OTT",
    "ANA", "ANA", "ANA", "ANA",
    "STL", "STL", "STL",
    "NSH", "NSH", "NSH",
    "CHI", "CHI", "CHI",
    "VAN",
    "MIN", "MIN"))

round4 <- data_frame(season="20142015",
  awayteam=c(
    "ANA", "STL", "NSH", "CHI", "VAN", "MIN", "WPG", "CGY",
    "ANA", "STL", "NSH", "CHI", "VAN", "MIN", "WPG", "CGY",
    "T.B", "WSH", "NYI", "DET", "OTT", "PIT",
    "T.B", "WSH", "NYI", "DET", "OTT", "PIT",
    "NSH", "CHI", "VAN", "MIN", "WPG", "CGY",
    "WSH", "NYI", "DET", "OTT", "PIT",
    "WSH", "NYI", "DET", "OTT", "PIT",
    "WSH", "NYI", "DET", "OTT", "PIT",
    "MIN", "WPG", "CGY",
    "MIN", "WPG", "CGY",
    "DET", "OTT", "PIT",
    "WPG", "CGY",
    "WPG", "CGY",
    "PIT",
    "CGY"),
  hometeam=c(
    "NYR", "NYR", "NYR", "NYR", "NYR", "NYR", "NYR", "NYR",
    "MTL", "MTL", "MTL", "MTL", "MTL", "MTL", "MTL", "MTL",
    "ANA", "ANA", "ANA", "ANA", "ANA", "ANA",
    "STL", "STL", "STL", "STL", "STL", "STL",
    "T.B", "T.B", "T.B", "T.B", "T.B", "T.B",
    "NSH", "NSH", "NSH", "NSH", "NSH",
    "CHI", "CHI", "CHI", "CHI", "CHI",
    "VAN", "VAN", "VAN", "VAN", "VAN",
    "WSH", "WSH", "WSH",
    "NYI", "NYI", "NYI",
    "MIN", "MIN", "MIN",
    "DET", "DET",
    "OTT", "OTT",
    "WPG",
    "PIT"))
```






Predict.

```{r}
round1$winner <- winners(round1, models, gamestats)
round1

round2 <- round2 %>%
  filter(awayteam %in% round1$winner, hometeam %in% round1$winner)
round2$winner <- winners(round2, models, gamestats)
round2

round3 <- round3 %>%
  filter(awayteam %in% round2$winner, hometeam %in% round2$winner)
round3$winner <- winners(round3, models, gamestats)
round3

round4 <- round4 %>%
  filter(awayteam %in% round3$winner, hometeam %in% round3$winner)
round4$winner <- winners(round4, models, gamestats)
round4
```

Next: [README](https://github.com/ilarischeinin/stanley)  
Previous: [4. Train models](model.html)
