---
title: "NHL Stanley Cup Playoff Predictions"
author: "Ilari Scheinin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

Previous: [1. Scrape raw data](scrape.html)  
Next: [3. Train models](model.html)

2. Process data
---------------

This script reads the data files produced by nhlscrapr and generates a summary
of each team's performance in the regular season each year.

First, load packages and the Canadian programming library:

```{r, echo=FALSE}
options(width=120)
```

```{r}
suppressMessages(library(dplyr))
source("canadian.R")
```

Then define a function to calculate each team's performance for a given season.

```{r}
process_season <- function(theseason) {
  load(file.path("source-data", paste0("nhlscrapr-", theseason, ".RData")))
  grand.data <- tbl_df(grand.data)

  gamestats <- grand.data %>%
    filter(ev.team %in% unique(grand.data$hometeam)) %>%
    filter(substring(gcode, 1, 1) == "2") %>%
    mutate(season==season) %>%
    group_by(season, gcode) %>%
    summarise(
      awayteam=first(awayteam),
      hometeam=first(hometeam),
      goals=sum(etype == "GOAL"),
      awaygoals=sum(ev.team == awayteam & etype == "GOAL") / goals,
      homegoals=sum(ev.team == hometeam & etype == "GOAL") / goals,
      shots=sum(etype %in% c("SHOT", "GOAL")),
      awayshots=sum(ev.team == awayteam & etype %in% c("SHOT", "GOAL")) / shots,
      homeshots=sum(ev.team == hometeam & etype %in% c("SHOT", "GOAL")) / shots,
      faceoffs=sum(etype == "FAC"),
      awayfaceoffs=sum(ev.team == awayteam & etype == "FAC") / faceoffs,
      homefaceoffs=sum(ev.team == hometeam & etype == "FAC") / faceoffs,
      penalties=sum(etype == "PENL"),
      awaypenalties=sum(ev.team == awayteam & etype == "PENL") / penalties,
      homepenalties=sum(ev.team == hometeam & etype == "PENL") / penalties,
      awaypp=(
        sum(ev.team == awayteam & etype == "GOAL" &
          away.skaters > home.skaters) -
        sum(ev.team == hometeam & etype == "GOAL" &
          away.skaters > home.skaters)) /
        homepenalties / penalties,
      homepp=(
        sum(ev.team == hometeam & etype == "GOAL" &
          away.skaters < home.skaters) -
        sum(ev.team == awayteam & etype == "GOAL" &
          away.skaters < home.skaters)) /
        awaypenalties / penalties,
      awaypk=1 - homepp,
      homepk=1 - awaypp)

  gamestats[na.eh(gamestats)] <- 0

  # the values in away.skaters and home.skaters sometimes imply pp/sh goals
  # when there has been zero penalties
  gamestats$awaypp[infinite.eh(gamestats$awaypp)] <- 0
  gamestats$homepp[infinite.eh(gamestats$homepp)] <- 0
  gamestats$awaypk[infinite.eh(gamestats$awaypk)] <- 0
  gamestats$homepk[infinite.eh(gamestats$homepk)] <- 0
  gamestats$season <- as.character(gamestats$season)
  gamestats <- gamestats %>% select(-goals, -shots, -faceoffs, -penalties)
  
  awaystats <- gamestats
  colnames(awaystats) <- sub("^away(.*)$", "\\1for", colnames(awaystats))
  colnames(awaystats) <- sub("^home(.*)$", "\\1against", colnames(awaystats))
  homestats <- gamestats
  colnames(homestats) <- sub("^home(.*)$", "\\1for", colnames(homestats))
  colnames(homestats) <- sub("^away(.*)$", "\\1against", colnames(homestats))
  homestats <- homestats[, colnames(awaystats)]

  # separate
  # awaystats <- awaystats %>%
  #   select(-gcode, -ends_with("against")) %>%
  #   rename(team=teamfor) %>%
  #   # ungroup() %>%
  #   group_by(season, team) %>%
  #   summarise_each(funs(mean))
  # colnames(awaystats) <- sub("for$", "", colnames(awaystats))
  # homestats <- homestats %>%
  #   select(-gcode, -ends_with("against")) %>%
  #   rename(team=teamfor) %>%
  #   # ungroup() %>%
  #   group_by(season, team) %>%
  #   summarise_each(funs(mean))
  # colnames(homestats) <- sub("for$", "", colnames(homestats))
  # return(list(awaystats=awaystats, homestats=homestats))

  # combined
  combined <- bind_rows(awaystats, homestats)
  combined <- combined %>%
    select(-gcode, -ends_with("against")) %>%
    rename(team=teamfor) %>%
    # ungroup() %>%
    group_by(season, team) %>%
    summarise_each(funs(mean))
  colnames(combined) <- sub("for$", "", colnames(combined))
  combined
}
```

Load all available seasons and run the function on each one, combine, and save.

```{r}
load(file.path("source-data", "nhlscrapr-core.RData"))
seasons <- unique(games$season)
rm(list=c("games", "roster.master", "roster.unique"))

# separate
# gamestats <- lapply(seasons, process_season)
# awaystats <- bind_rows(lapply(gamestats, "[[", "awaystats"))
# homestats <- bind_rows(lapply(gamestats, "[[", "homestats"))

# combined
gamestats <- bind_rows(lapply(seasons, process_season))
gamestats

saveRDS(gamestats, "preprocessed.rds")
```

```{r, results='asis'}
knitr::kable(gamestats[1:10, ], digits=3)
```

Next: [3. Train models](model.html)  
Previous: [1. Scrape raw data](scrape.html)
