---
title: "NHL Stanley Cup Playoff Predictions"
author: "Ilari Scheinin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

Previous: [1. Canadian programming](canadian.html)  
Next: 3. Process data

1. Scrape raw data
------------------

To scrape the data from [nhl.com](http://www.nhl.com), I am using the
[nhlscrapr](https://github.com/acthomasca/nhlscrapr)
[package](http://cran.r-project.org/web/packages/nhlscrapr/index.html). It has
a single command `compile.all.games()`, which downloads and compiles
everything together.

However, it waits 20 seconds between every game, and therefore takes more than
3.5 days to run. Instead, one might want to use something like the two options
below to download games one-by-one or by season, and to set a shorter time
interval.

```{r}
suppressMessages({
  library(nhlscrapr)
})

compile.all.games()
```

To download games one-by-one or by season.

```{r, eval=FALSE}

# get full list of games available
games <- full.game.database()

# download by game
apply(games, 1, function(game) {
  download.single.game(season=game["season"], gcode=game["gcode"], wait=2)
  gc()
})

# download by season
lapply(unique(games$season), function(season) {
  download.games(games[games$season == season, ], wait=2)
  gc()
})

# and once downloaded, compile everything together
compile.all.games()
```

Next: 3. Process data  
Previous: [1. Canadian programming](canadian.html)
